\documentclass[11pt, english, fleqn, DIV=15, headinclude, BCOR=1.5cm]{scrartcl}

\usepackage[
    bibatend,
    color,
]{../header}

\usepackage{tikz}
\usepackage{pdflscape}

\usepackage[tikz]{mdframed}
\newmdtheoremenv[%
    backgroundcolor=black!5,
    innertopmargin=\topskip,
    splittopskip=\topskip,
]{theorem}{Theorem}[section]

\hypersetup{
    pdftitle=
}

\newcounter{totalpoints}
\newcommand\punkte[1]{#1\addtocounter{totalpoints}{#1}}

\newcounter{problemset}
\setcounter{problemset}{5}

\subject{physics606 -- Advanced Quantum Theory}
\ihead{physics606 -- Problem Set \arabic{problemset}}

\title{Problem Set \arabic{problemset}}

\publishers{Group 2 -- Dilege Gülmez}
\ofoot{Group 2 -- Dilege Gülmez}

\newmdenv[%
    backgroundcolor=black!5,
    frametitlebackgroundcolor=black!10,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    splittopskip=\topskip,
    frametitle={Problem statement},
    frametitlerule=true,
    nobreak=true,
]{problem}

\newmdenv[%
    backgroundcolor=white,
    frametitlebackgroundcolor=black!10,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    innerbottommargin=8cm,
    splittopskip=\topskip,
    frametitle={Side question},
    frametitlerule=true,
]{question}


\author{
    Martin Ueding \\ \small{\href{mailto:mu@martin-ueding.de}{mu@martin-ueding.de}}
    \and
    Lino Lemmer \\ \small{\href{mailto:l2@uni-bonn.de}{l2@uni-bonn.de}}
}
\ifoot{Martin Ueding, Lino Lemmer}

\ohead{\rightmark}

\begin{document}

\maketitle

\vspace{3ex}

\begin{center}
    \begin{tabular}{rrr}
        problem number & achieved points & possible points \\
        \midrule
        1 & & \punkte{11} \\
        2 & & \punkte{10} \\
        3 & & \punkte{13} \\
        \midrule
        total & & \arabic{totalpoints}
    \end{tabular}
\end{center}

\section{Exponentiating operators}

\begin{problem}
    The exponential of an operator is defined as
    \[
        \eup^A = \sum_{n=0}^\infty \frac{A^n}{n!}.
    \]
\end{problem}

\subsection{Inverse}

\begin{problem}
    Show that $\eup^A \eup^{-A} = 1$ using only the definition.
\end{problem}

We start with the definition
\begin{align*}
    \eup^A \eup^{-A}
    &= \sum_{n=0}^\infty \frac{A^n}{n!} \sum_{m=0}^\infty \frac{[-A]^m}{m!} \\
    \intertext{%
        and move the terms.
    }
    &= \sum_{n=0}^\infty \sum_{m=0}^\infty \frac{A^n}{n!} \frac{[-A]^m}{m!} \\
    &= \sum_{n=0}^\infty \sum_{m=0}^\infty \frac{[-1]^m A^{m+n}}{m!n!} \\
    \intertext{%
        We now think about those two sums as of a lattice where each point can
        be labelled using $(m, n)$. Currently, the summation is done in rows
        and columns. We change this to a diagonal iterations. A new parameter
        $a := m + n$ is introduced to change the summation to:
    }
    &= \sum_{a=0}^\infty \sum_{m=0}^a \frac{[-1]^m A^a}{m![a-m]!}. \\
    \intertext{%
        This is somewhat related to the argument to show that $\Q$ is still
        countable infinite. The case $a = 0$ is special and will be taken out
        of the sum, it will be just 1. We now rearrange the terms bit more and
        introduce $a!$ which cancels in total.
    }
    &= 1 + \sum_{a=1}^\infty \frac{A^a}{a!} \sum_{m=0}^a \frac{a!}{m![a-m]!}
    [-1]^m \\
    \intertext{%
        The large fraction is just the binomial coefficient. We also add a
        $1^{a-m}$.
    }
    &= 1 + \sum_{a=1}^\infty \frac{A^a}{a!} \sum_{m=0}^a \binom{a}{m} [-1]^m
    1^{a-m} \\
    \intertext{%
        Using the binomial theorem, we can write this more compact as
    }
    &= 1 + \sum_{a=1}^\infty \frac{A^a}{a!} \underbrace{[-1 + 1]^a}_{=0} \\
    \intertext{%
        which means that the second part vanishes for any $a$. Here it is clear
        that including $a = 0$ in the sum would give the indeterminate
        expression $0^0$ which would require extra treatment either way. We are
        left with
    }
    &= 1.
\end{align*}

\subsection{Sum in exponent}

\begin{problem}
    Show that
    \[
        \eup^{A+B} = \eup^A \eup^B \eup^{[A,B]}
    \]
    if the commutator $[A,B] =: c$ is a complex number.
\end{problem}

\subsection{Commutation of exponentials}

\begin{problem}
    Show that
    \[
        \eup^A \eup^B = \eup^B \eup^A \eup^{[A,B]}.
    \]
\end{problem}

In the previous problem we got
\[
    \eup^{A + B} = \eup^A \eup^B \eup^{-c/2}
\]
where $c := [A, B] \in \C$. We can just exchange $A$ and $B$ and obtain
\[
    \eup^{B + A} = \eup^B \eup^A \eup^{c/2}.
\]

Using the first equation, we can move the $\eup^{-c/2}$ to the other side and
get
\begin{align*}
    \eup^A \eup^B &= \eup^{A + B} \eup^{c/2} \\
    \intertext{%
        There, we can commute $A$ and $B$ in the \emph{sum} in the exponent.
    }
    \iff \eup^A \eup^B &= \eup^{B + A} \eup^{c/2} \\
    \intertext{%
        Using the equation we obtained from exchanging $A$ and $B$, we yield
    }
    &= \eup^B \eup^A \eup^{c/2} \eup^{c/2} \\
    &= \eup^B \eup^A \eup^{c},
\end{align*}
which is the desired result.

\subsection{Generalized commutator}

\begin{problem}
    Show by induction that
    \[
        [A, B]_n = \sum_{i=0}^n \frac{n!}{n![n-i]!} A^{n-i} B [-A]^i.
    \]
\end{problem}

The way the problem is given, a $n!$ could be canceled. It can be seen for $n =
2$ that the relation given does not hold. The left side is
\[
    [A, B]_2 = A^2 B - 2 ABA + BA^2.
\]
The right side with $n!/n!$ canceled gives
\[
    \sum_{i=0}^2 \frac{1}{[2-i]!} A^{2-i} B [-A]^i
    = \frac12 A^2 B - ABA + B^2.
\]
The powers are right, just the prefactors are off.

We think that this looks so close to the binomial coefficient which was already
used much in the previous problems that it probably should be $i!$ in the
denominator. Using $i!$ in the denominator, we can proof the identity using
induction. Therefore, we assume the following relation:
\[
    [A, B]_n
    = \sum_{i=0}^n \frac{n!}{i![n-i]!} A^{n-i} B [-A]^i
    = \sum_{i=0}^n \binom ni \, A^{n-i} B [-A]^i.
\]

\begin{proof}
    We show that the relation holds for $n = 0$:
    \[
        [A, B]_0 = B
        \eqnsep
        \sum_{i=0}^0 \binom00 \, A^0 B [-A]^0 = B.
    \]
    This holds. Next is the induction step.
    \begin{align*}
        [A, B]_{n+1}
        &= [A, [A,B]_n] \\
        &= A [A,B]_n - [A,B]_n A \\
        \intertext{%
            We use the relation that we want to show, this is allowed in
            induction.
        }
        &= A \sum_{i=0}^n \binom{n}{i} \, A^{n-i} B [-A]^i - \sum_{i=0}^n
        \binom{n}{i} A^{n-i} B [-A]^i A \\
        \intertext{%
            We factor the additional factors $A$ into the expression.
        }
        &= \sum_{i=0}^n \binom{n}{i} \, A^{n+1-i} B [-A]^i + \sum_{i=0}^n
        \binom{n}{i} A^{n-i} B [-A]^{i+1} \\
        \intertext{%
            Now we can shift the indices a bit. We introduce
            \[
                j := i + 1
                \eqnsep
                i = j - 1
            \]
            and put that into the expression.
        }
        &= \sum_{i=0}^n \binom{n}{i} \, A^{n+1-i} B [-A]^i + \sum_{j=1}^{n+1}
        \binom{n}{j-1} \, A^{n+1-j} B [-A]^{j} \\
        \intertext{%
            We branch off the first and last summand of the first and last sum,
            \emph{respectively}. Then we combine the middle part.
        }
        &= A^{n+1} B +  \sum_{i=1}^n \sbr{\binom{n}{i} + \binom{n}{i-1}}
        A^{n+1-i} B [-A]^i + B [-A]^{n+1} \\
        \intertext{%
            There is a handy addition theorem for binomial coefficients which
            we will use for the bracket.
        }
        &= A^{n+1} B +  \sum_{i=1}^n \binom{n+1}{i} \, A^{n+1-i} B [-A]^i
        + B [-A]^{n+1} \\
        \intertext{%
            We add ones to the stray summands.
        }
        &= \binom{n+1}{0} A^{n+1} B + \sum_{i=1}^n \binom{n+1}{i} \, A^{n+1-i} B [-A]^i
        + \binom{n+1}{n+1} B [-A]^{n+1} \\
        \intertext{%
            Now we can combine it into a single sum.
        }
        &= \sum_{i=0}^{n+1} \binom{n+1}{i} \, A^{n+1-i} B [-A]^i
    \end{align*}
    That is exactly what has to be shown. Therefore, this relation holds for
    any $n$.
\end{proof}

Now that we got this identity at our hands, we can aim for the actual task at
hand:

\begin{problem}
    Show that
    \[
        \eup^A B \eup^{-A} = \sum_{n = 0}^\infty \frac{1}{n!} [A, B]_n.
    \]
\end{problem}

We start with the left side of this equation:
\begin{align*}
    \eup^A B \eup^{-A}
    &= \sum_{d=0}^\infty \frac{A^d}{d!} B \sum_{f=0}^\infty \frac{[-A]^f}{f!}.
    \\
    \intertext{%
        We interchange the summations again and group the terms.
    }
    &= \sum_{d=0}^\infty \sum_{f=0}^\infty \frac{1}{f!d!} A^d B [-A]^f \\
    \intertext{%
        By now, the change in the lattice traversal was applied many times. We
        introduce
        \[
            i := f
            \eqnsep
            n := d + f
        \]
        and get
    }
    &= \sum_{n=0}^\infty \sum_{i=0}^n \frac{1}{i![n-i]!} A^d B [-A]^f. \\
    \intertext{%
        Now we introduce a factor of $n!$.
    }
    &= \sum_{n=0}^\infty \frac{1}{n!} \sum_{i=0}^n \frac{n!}{i![n-i]!} A^d B [-A]^f \\
    \intertext{%
        Using the identity we proved at the beginning of this exercise, we can
        write this as
    }
    &= \sum_{n=0}^\infty \frac{1}{n!} [A, B]_n
\end{align*}
which is what we wanted to show.

\section{First order time dependent perturbation theory}

\newcommand\0{^{(0)}}

\subsection{Argumentation}

\begin{problem}
    Argue that $P_{fi}(t) = |c_f(t)|^2$.
\end{problem}

The probability that the system will transition from $\ket{i\0}$ to $\ket{f\0}$
in the time interval $[t_0, t]$ is given by the modulus squared of the overlap
intergral:
\begin{align*}
    P_{fi}(t)
    &= |\braket{f\0 | \psi(t)}|^2. \\
    \intertext{%
        We can insert the definition of $\psi(t)$ and get
    }
    &= \left| \sum_n c_n(t) \braket{f\0 | n\0} \right|^2. \\
    \intertext{%
        The scalar product gives a $\delta_{fn}$, reducing the sum to
    }
    &= |c_f(t)|^2.
\end{align*}

\subsection{Relation}

\begin{problem}
    What is the relation between $d_n(t)$ and $c_n(t)$?
\end{problem}

We have
\[
    \ket{\psi(t)} = \sum_n d_n(t) \ket{\psi_n(t)}.
\]
The states $\ket{\psi_n(t)}$ are defined as the unperturbed time evolution of
the basis states $\ket{n\0}$. Using the time evolution operator, we can write
this as
\[
    \ket{\psi_n(t)} = U\0(t-t_0) \ket{n\0}.
\]
Now we insert this into the above relation:
\[
    \ket{\psi(t)} = \sum_n d_n(t) U\0(t-t_0) \ket{n\0}.
\]
By comparison, we get
\[
    c_n(t) = d_n(t) U\0(t-t_0) = d_n(t) \exp\del{-\frac\iup\hbar E_n\0
    [t-t_0]}.
\]

\subsection{Time derivative of coefficient}

\begin{problem}
    Show that
    \[
        \iup \hbar \dot d_f = \sum_n \braket{f\0 | H_1(t) | n\0} \exp(\iup
        \omega_{fn} [t-t_0]) d_n(t).
    \]
\end{problem}

We have
\[
    \ket{\psi(t)} = \sum_n d_n(t) \ket{\psi_n(t)}.
\]
We can apply $\bra{\psi_m(t)}$ to both sides and get
\begin{align*}
    \braket{\psi_m(t) | \psi(t)}
    &= \sum_n d_n(t) \braket{\psi_m(t) | \psi_n(t)}. \\
    \intertext{%
        We include the time evolution operator explicitly:
    }
    &= \sum_n d_n(t) \braket{m\0 | U\0(t-t_0)^\dagger U\0(t-t_0) | n\0}. \\
    \intertext{%
        However, that just cancels out because the time evolution operator is
        unitary. We are left with
    }
    &= \sum_n d_n(t) \braket{m\0 | n\0} \\
    \intertext{%
        which reduces to
    }
    &= d_m(t)
\end{align*}
since the unperturbed eigenstates are orthonormal. As we have shown here, that
also holds for any time $t$ in general. Now we have isolated an expression for
$d_n(t)$ where we can take the time derivative. We start with the derived
expression:
\begin{align*}
    \iup \hbar \dot d_f(t)
    &= \iup \hbar \od{}t \braket{\psi_f(t) | \psi(t)}. \\
    \intertext{%
        Here we use the Schrödinger equation for both the states. This looks a
        bit clumsy, but we have to use the product rule here since both states
        depend on time. We have
    }
    &= \sbr{\bra{\psi_f(t)} H^\dagger} \ket{\psi(t)}
    + \bra{\psi_f(t)} \sbr{H \ket{\psi(t)}}. \\
    \intertext{%
        With $H$ being a hermitian operator, $H = H^\dagger$ holds. This means
        that we end up with
    }
    &= 2 \braket{\psi_f(t) | H | \psi(t)}. \\
    \intertext{%
        We now insert the expansion of the state $\ket{\psi(t)}$ and obtain
    }
    &= 2 \sum_n d_n(t) \braket{\psi_f(t) | H | \psi_n(t)}. \\
    \intertext{%
        We expand the Hamiltonian into its parts, which are
    }
    &= 2 \sum_n d_n(t) \braket{\psi_f(t) | H_0 + H_1(t) | \psi_n(t)}. \\
    \intertext{%
        Using the linearity of the scalar product, we can expand this into two
        parts.
    }
    &= 2 \sum_n d_n(t) \sbr{
        \braket{\psi_f(t) | H_0 | \psi_n(t)}
        + \braket{\psi_f(t) | H_1(t) | \psi_n(t)}
    } \\
    \intertext{%
        The energies of the unperturbed eigenstates are still $E_n\0$, so we
        can just insert them there
    }
    &= 2 \sum_n d_n(t) \sbr{
        E_n\0 \delta_{fn}
        + \braket{\psi_f(t) | H_1(t) | \psi_n(t)}
    } \\
    \intertext{%
        and get rid of the sum for that summand.
    }
    &= 2 d_f(t) E_f\0
    + 2 \sum_n d_n(t) \braket{\psi_f(t) | H_1(t) | \psi_n(t)} \\
    \intertext{%
        We again write the time dependent basis states with the time evolution
        operator.
    }
    &= 2 d_f(t) E_f\0
    + 2 \sum_n d_n(t) \braket{f\0 | U\0(t-t_0)^\dagger H_1(t) U\0(t-t_0) | n\0} \\
    \intertext{%
        Next we insert the time evolution operator explicitly for the states
        left and right. We obtain
    }
    &= 2 d_f(t) E_f\0 \\
    &\quad+ 2 \sum_n d_n(t) \Braket{f\0 | \exp\del{\frac\iup\hbar E_f\0 [t-t_0]}
    H_1(t) \exp\del{-\frac\iup\hbar E_f\0 [t-t_0]} | n\0} \\
    \intertext{%
        The time evolution operator is just a scalar now and can be moved out
        of the scalar product. We combine the two exponential functions and use
        the $\omega_{fn}$ that is given on the problem set as well.
    }
    &= 2 d_f(t) E_f\0 + 2 \sum_n \braket{f\0 | H_1(t) | n\0} \exp\del{\iup \omega_{fn}[t-t_0]} d_n(t)
\end{align*}

Apart from an additional term and a factor of 2, this is the term we end up
with.

\begin{question}
    What do we need to do differently to get rid of that additional summand and
    the factor of 2?
\end{question}

\subsection{First order expression}

\section{Perturbed harmonic oscillator}


\end{document}

% vim: spell spelllang=en tw=79
