\documentclass[11pt, english, fleqn, DIV=15, headinclude, BCOR=1.5cm]{scrartcl}

\usepackage[
    bibatend,
    color,
]{../header}

\usepackage{tikz}
\usepackage{pdflscape}

\usepackage[tikz]{mdframed}
\newmdtheoremenv[%
    backgroundcolor=black!5,
    innertopmargin=\topskip,
    splittopskip=\topskip,
]{theorem}{Theorem}[section]

\hypersetup{
    pdftitle=
}

\newcounter{totalpoints}
\newcommand\punkte[1]{#1\addtocounter{totalpoints}{#1}}

\newcounter{problemset}
\setcounter{problemset}{10}

\subject{physics606 -- Advanced Quantum Theory}
\ihead{physics606 -- Problem Set \arabic{problemset}}

\title{Problem Set \arabic{problemset}}

\publishers{Group 2 -- Dilege Gülmez}
\ofoot{Group 2 -- Dilege Gülmez}

\newmdenv[%
    backgroundcolor=black!0,
    frametitlebackgroundcolor=black!0,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    splittopskip=\topskip,
    frametitle={Problem statement},
    frametitlerule=true,
    nobreak=true,
]{problem}

\newmdenv[%
    backgroundcolor=white,
    frametitlebackgroundcolor=black!0,
    roundcorner=5pt,
    skipabove=\topskip,
    innertopmargin=\topskip,
    innerbottommargin=8cm,
    splittopskip=\topskip,
    frametitle={Side question},
    frametitlerule=true,
]{question}

\newcommand\an{^\text{angular}}
\newcommand\ra{^\text{radial}}


\author{
    Martin Ueding \\ \small{\href{mailto:mu@martin-ueding.de}{mu@martin-ueding.de}}
    \and
    Lino Lemmer \\ \small{\href{mailto:l2@uni-bonn.de}{l2@uni-bonn.de}}
}
\ifoot{Martin Ueding, Lino Lemmer}

\ohead{\rightmark}

\begin{document}

\maketitle

\vspace{3ex}

\begin{center}
    \begin{tabular}{rrr}
        problem number & achieved points & possible points \\
        \midrule
        1 & & \punkte{15} \\
        2 & & \punkte{13} \\
        \midrule
        total & & \arabic{totalpoints}
    \end{tabular}
\end{center}

\section{Scattering on a hard sphere}

\subsection{Phase shift}

\begin{problem}
    Write down the radial wave function.
\end{problem}

Inside the sphere, the potential is infinite, so it is exactly zero for any
angular momentum $l$:
\[
    R_l^<(r) = 0.
\]
The part outside the sphere is independent of the potential,
\[
    R_l^>(r) = \frac12 \sbr{h_l^*(kr) + \eup^{2\iup \delta_l} h_l(kr)},
\]
where $h_l := h_l^{(1)}$ \parencite[(18.59)]{Schwabl/Quantenmechanik}.

\begin{problem}
    Use the continuity of the wave function at $r = r_0$ together with the
    asymptotic expression for the wave function at $r \to \infty$ to show that
    \[
        \tan(\delta_l) = \frac{j_l(kr_0)}{n_l(kr_0)}.
    \]
\end{problem}

At the boundary $r_0$ the wave function $\psi$ has to be both continuous and
differentiable. Both criteria can be checked at the same time by the virtue of
the logarithmic derivative, which is defined in
\parencite[(18.60)]{Schwabl/Quantenmechanik} as
\[
    \alpha_l := \eval{\od{\,\log\del{R_l^<(r)}}r}_{r = r_0}
    = \eval{\frac{1}{R_l^<(r)} \od{R_l^<(r)}r}_{r = r_0}.
\]
As mentioned before, the wave function $\psi$ and its radial part $R$ will be
zero within this sphere of radius $r_0$. Also, since the potential makes an
infinite jump, $R$ will not be differentiable, just like the energy
eigenfunctions of the infinite potential well. Both factors in the logarithmic
derivative will be infinite, such that $\alpha_l = \infty$ is result. The
logarithmic derivative of the inner and outer solutions have to match up. This
leads to \parencite[(18.61)]{Schwabl/Quantenmechanik}:
\[
    \frac{1}{R_l^>(r)} \od{R_l^>(r)}r \overset != \alpha_l.
\]
The spherical Hankel functions can be split up into spherical Bessel $j_l$ and
spherical von~Neumann $n_l$ functions. Then the previous equation can be
transformed into \parencite[(18.62)]{Schwabl/Quantenmechanik}:
\[
    \eup^{2 \iup \delta_l} -1 = 2 \eval{\frac{\od{j_l}r - \alpha_l j_l}{\alpha_l h_l
    - \od{h_l}r}}_{r=r_0}.
\]
Solving for $\delta_l$ gives \parencite[(18.62)]{Schwabl/Quantenmechanik}:
\[
    \tan(\delta_l) = \eval{\frac{\od{j_l}r - \alpha_l j_l}{\od{n_l}r - \alpha_l
    n_l}}_{r=r_0}.
\]
Since $\alpha_l$ is infinite, the remainder of this fraction is
\[
    \tan(\delta_l) = \frac{j_l(kr_0)}{n_l(kr_0)},
\]
which is also the result mentioned for the hard sphere in
\parencite[(18.63)]{Schwabl/Quantenmechanik}.

\subsection{Negative shift}

\begin{problem}
    Use the explicit expressions for $j_0(x)$ and $n_0(x)$ to show that
    $\delta_0 = - kr_0$.
\end{problem}

The first order of those spherical functions are given by
\parencite{MathWorld/BesselFunctionoftheSecondKind,MathWorld/BesselFunctionoftheFirstKind}:
\[
    j_0(x) = \frac{\sin(x)}{x}
    \eqnsep
    n_0(x) = - \frac{\cos(x)}{x}.
\]
Inserting this into the previous expression for the phase shift gives us
\[
    \tan(\delta_l) = - \frac{\sin(kr_0)}{\cos(kr_0)}
\]
which simplifies to $\delta_l = - kr_0$. The minus sign can be put into both
sine and cosine, then the tangent can be inverted.

\begin{problem}
    Why is the phase shift negative here?
\end{problem}

The shift is negative since the wave is scattered on a repulsive potential.

\subsection{Expansion for small arguments}

\begin{problem}
    Use the expansions of $j_l$ and $n_l$ at small argument to show that
    $\delta_l \propto [kr_0]^{2l+1}$, just as for scattering on a spherical
    potential well.
\end{problem}

In the ninth exercise, we derived this approximation, it was also given as a
control result:
\[
    j_l(x) = x^l \frac{2^l l!}{[2l+1]!} + \mathcal O(x^{l+2}).
\]
The von~Neumann functions have a similar representation using repeated
differentiation \parencite{wikipedia/bessel_dgl}:
\[
    j_l(x) = [-x]^l \sbr{\frac 1x \od{}x}^l \frac{\sin(x)}{x}
    \eqnsep
    n_l(x) = - [-x]^l \sbr{\frac 1x \od{}x}^l \frac{\cos(x)}{x}.
\]

The derivation is similar. We expand the cosine in terms of its power series:
\begin{align*}
    n_l(x)
    &= - \sum_{n=0}^l \frac{[-1]^n}{[2n]!} [-x]^l \sbr{\frac 1x \od{}x}^l
    x^{2n-1}.
    \intertext{%
        Now we drop all but the lowest power, which comes at $n = 0$. This is
        the same as just using the first order expansion of the cosine.
    }
    &\approx - [-x]^l \sbr{\frac 1x \od{}x}^l x^{-1}.
    \intertext{%
        Assuming that $l \geq 1$, we can perform the derivatives and obtain
    }
    &= - [2l-1]!! \, x^{-1-l}.
\end{align*}

Then we simply have
\[
    \frac{j_l(x)}{n_l(x)} \propto \frac{x^l}{x^{-1-l}} = x^{2l+1}.
\]
However, this is $\tan(\delta_l)$, not $\delta_l$ itself. Do we make the
approximation $\tan(x) = x + \mathcal O(x^3)$ here as well?

\subsection{Asymptotic expansion}

\begin{problem}
    Use the asymptotic expressions for $j_l$ and $n_l$ for large arguments to
    show that
    \[
        \lim_{k r_0 \to \infty} \delta_l(k) = - kr_0 + \frac{l \piup}2.
    \]
\end{problem}

\parencite{wikipedia/bessel_dgl} gives the following large argument expressions
for the non-spherical functions:
\[
    J_\nu(x) = \sqrt{\frac{2}{\piup x}} \cos\del{x - \frac{\nu\piup}2 -
    \frac\piup4}
    \eqnsep
    N_\nu(x) = \sqrt{\frac{2}{\piup x}} \sin\del{x - \frac{\nu\piup}2 -
    \frac\piup4}.
\]
The relation to the spherical ones is given by
\[
    j_l(x) = \sqrt{\frac\piup{2x}} J_{l + 1/2}(x)
    \eqnsep
    n_l(x) = \sqrt{\frac\piup{2x}} N_{l + 1/2}(x).
\]
We now plug this together and get
\[
    j_l(x) = \frac1x \cos\del{x - \frac{l\piup}2}
    \eqnsep
    n_l(x) = \frac1x \sin\del{x - \frac{l\piup}2}.
\]
The ratio gives us
\[
    \tan(\delta_l) = \tan\del{kr_0 - \frac{l\piup}2}
    \iff
    \delta_l = kr_0 - \frac{l\piup}2.
\]
The problem is that the sign is wrong.

\subsection{Total cross section}

We start with equation \parencite[(18.40)]{Schwabl/Quantenmechanik}:
\begin{align*}
    \sigma
    &= \frac{4\piup}{k^2} \sum_{l=0}^{kr_0} [2l+1] \sin(\delta_l)^2. \\
    \intertext{%
        Now we insert the result of $\delta_l$ from the previous part.
    }
    &= \frac{4\piup}{k^2} \sum_{l=0}^{kr_0} [2l+1] \sin\del{-kr_0 +
    \frac{l\piup}2}^2
    \intertext{%
        We now split the sum into even and odd values of $l$, introducing $m$.
    }
    &= \frac{4\piup}{k^2} \sum_{m=0}^{kr_0/2} \sbr{
        [2[2m]+1] \sin\del{-kr_0 + \frac{2m\piup}2}^2
        + [2[2m+1]+1] \sin\del{-kr_0 + \frac{[2m+1]\piup}2}^2
    } \\
    &= \frac{4\piup}{k^2} \sum_{m=0}^{kr_0/2} \sbr{
        [4m+1] \sin\del{-kr_0 + m\piup}^2
    + [4m+3] \sin\del{-kr_0 + \frac{[2m+1]}2 \piup}^2
    }
    \intertext{%
        The first sine will always be a sine for any value of $m$. The square
        will take care of the sign, so we can just write this as a pure sine of
        the first summand. The second sine will always be a cosine.
    }
    &= \frac{4\piup}{k^2} \sum_{m=0}^{kr_0/2} \sbr{
        [4m+1] \sin(kr_0)^2 + [4m+3] \cos(kr_0)^2
    }
    \intertext{%
        We combine sine and cosine.
    }
    &= \frac{4\piup}{k^2} \sum_{m=0}^{kr_0/2} [8m+4]
    \intertext{%
        The sum can now be evaluated and gives
    }
    &= \frac{4\piup}{k^2} \sbr{[kr_0]^2 + 4kr_0 + 4}.
    \intertext{%
        We simplify more and only keep the leading term in $k$ since $k$ is
        supposed to be large.
    }
    &= 4 \piup r_0^2
\end{align*}
So this is a factor 4 larger than the geometrical cross section.

\section{Scattering length}

\subsection{Relation to scattering length}

We start with equation~(3) from the problem set:
\begin{align*}
    \tan(\delta_0) &= - k r_0 \frac{\alpha_0}{1 + \alpha_0}. \\
    \intertext{%
        Then we invert this equation.
    }
    \iff \cot(\delta_0) &= - \frac{1}{k r_0 \frac{\alpha_0}{1 + \alpha_0}} \\
    \intertext{%
        Now we multiply with $k$ on both sides.
    }
    \iff k \cot(\delta_0) &= - \frac{1}{r_0 \frac{\alpha_0}{1 + \alpha_0}} \\
    \intertext{%
        As a last step, we identify $R_0 = r_0 \frac{\alpha_0}{1 + \alpha_0}$.
    }
    \iff k \cot(\delta_0) &= - \frac{1}{R_0}
\end{align*}

\subsection{}

\begin{problem}
    Show that
    \[
        \eup^{2 \iup \delta_0}
        = \frac{2 \iup}{\cot(\delta_0) - \iup}
        = \frac{2k R_0}{\iup - kR_0}
        = - 2 \iup k R_0 + \mathcal O(k^2).
    \]
\end{problem}

The correct version of the first equality is given in
\parencite[(18.95)]{Schwabl/Quantenmechanik} and has an additional term
compared to the one from the problem set\footnote{Regarding the version from
the evening of Wednesday, 2014-12-10.}:
\[
    \eup^{2 \iup \delta_0} \underbrace{-1}
    = \frac{2 \iup}{\cot(\delta_0) - \iup}.
\]
With that, this is rather straightforward. We define $w := \eup^{\iup
\delta_0}$ to write less stuff. The derivation is just the simplification
backwards:
\begin{align*}
    \exp(2 \iup \delta_0)
    &= w^2 - 1 \\
    \intertext{%
        Expand.
    }
    &= \frac{w^{-1} [w^2 -1]}{w^{-1}} \\
    \intertext{%
        Factor out.
    }
    &= \frac{w - w^{-1}}{w^{-1}} \\
    \intertext{%
        Introduce factor of 2.
    }
    &= 2 \frac{w - w^{-1}}{2 w^{-1}} \\
    \intertext{%
        Make this into a more complex double fraction.
    }
    &= 2 \frac{1}{\frac{2 w^{-1}}{w - w^{-1}}} \\
    \intertext{%
        Invent more terms.
    }
    &= \frac{2}{\frac{w - w^{-1} - w + w^{-1}}{w - w^{-1}}} \\
    \intertext{%
        Separate fractions.
    }
    &= \frac{2}{\frac{w + w^{-1}}{w - w^{-1}} - 1} \\
    \intertext{%
        Introduce a factor of $\iup$.
    }
    &= \frac{2\iup}{\iup\frac{w + w^{-1}}{w - w^{-1}} - \iup} \\
    \intertext{%
        Write said factor as another double fraction.
    }
    &= \frac{2\iup}{\frac{1/2}{1/[2\iup]}\frac{w + w^{-1}}{w - w^{-1}} - \iup} \\
    \intertext{%
        Recognize cosine and sine.
    }
    &= \frac{2\iup}{\frac{\cos(\delta_0)}{\sin(\delta_0) - \iup}} \\
    \intertext{%
        Compress this into a cotangent.
    }
    &= \frac{2\iup}{\cot(\delta_0) - \iup}
\end{align*}
That was the first part.

From here, the next part is a bit easier:
\begin{align*}
    \frac{2\iup}{\cot(\delta_0) - \iup}
    &= \frac{2\iup}{- \frac{1}{k R_0} - \iup} \\
    \intertext{%
        Combine the fractions.
    }
    &= - \frac{2\iup}{\frac{1 + \iup k R_0}{k R_0}} \\
    \intertext{%
        Simplify the double fraction.
    }
    &= - \frac{2\iup k R_0}{1 + \iup k R_0} \\
    \intertext{%
        Introduce a $-\iup$ in the denominator.
    }
    &= - \frac{2\iup k R_0}{[-\iup][\iup - k R_0]} \\
    \intertext{%
        Cancel $-\iup$.
    }
    &= \frac{2 k R_0}{\iup - k R_0}
\end{align*}
That is the second part.

The last part is a Taylor expansion around $k = 0$. The steps are:
\begin{gather*}
    f(k) := \frac{2 k R_0}{\iup - k R_0} \\ 
    f(0) = 0 \\
    f'(k) = \frac{2 R_0 [\iup - kR_0] - 2 k R_0[-R_0]}{[\iup - kR_0]^2} \\
    f'(0) = - 2 \iup R_0
\end{gather*}
So the approximation is
\[
    \frac{2 k R_0}{\iup - k R_0}
    = - 2 \iup k R_0 + \mathcal O(k^2).
\]

\subsection{$S$-wave amplitude}

Using the equation from \parencite[(18.96)]{Schwabl/Quantenmechanik},
\[
    f_0 = \frac{2kR_0}{\iup - k R_0} \frac{1}{2 \iup k},
\]
we find using the approximation that we have just derived that
\[
    f_0 \approx - R_0.
\]
Since the total cross section for $S$-wave scattering is given by
\[
    \sigma = \int \dif \Omega \, |f_0|^2,
\]
we obtain the desired result of
\[
    \sigma = 4 \piup R_0^2.
\]

\subsection{Wavefunction outside}

In this problem, a second $R_0$ comes up. Until now, $R_0 \in \R$ was the
scattering length. Now this second $R_0$ is an element of the mapping $\R \to
\R$. It stands for the radial part of the $S$-wave function. It is the special
case $l = 1$ of $R_l(r)$. As a rule of thumb, every $R_0(r)$ means the radial
wave function, $R_0$ without function arguments means the scattering length.
Luckily, this problem does not ask anything about the function $R_0$ itself, so
the notation would be impossible.

Anyway, to the problem at hand, I digress. We are given another form of the
wave function outside of the sphere. We had a version by
\citeauthor{Schwabl/Quantenmechanik}. This one here is
\begin{align*}
    R_0(r)
    &\propto \cos(\delta_0) j_0(kr) - \sin(\delta_0) n_0(kr). \\
    \intertext{%
        Now we factor out a factor of cosine and hide it in the
        proportionality since it does not depend on the radius $r$.
    }
    &\propto j_0(kr) - \tan(\delta_0) n_0(kr) \\
    \intertext{%
        Here we use equation~(4) from the problem set to replace the tangent.
        Keep in mind that this $R_0$ is not the same as the other one.
    }
    &= j_0(kr) + k R_0 n_0(kr) \\
    \intertext{%
        Next are the approximations for the Bessel and von~Neumann functions.
        For the zeroth order, they are quite simple, just $1$ and $-1/[kr]$.
    }
    &= 1 - \frac{R_0}r  \\
    \intertext{%
        We change the order of the summands and again use the proportionality
        here.
    }
    &\propto \frac{R_0}r - 1 \\
    \intertext{%
        Now we just multiply both sides with the radius $r$.
    }
    r R_0(r) &\propto R_0 - r
\end{align*}
So this gives us a proportionality between the two $R_0$.

\end{document}

% vim: spell spelllang=en tw=79
